Understanding the Lomb-Scargle Periodogram\\ {\it (Draft Version \today)}

### 2017-03-23 PracticalLombScargle.tex
minor edits and rewordings

We are generally not interested in the distribution of one particular
randomly-chosen frequency,
but rather the distribution of {\it the highest peak} of the
periodogram, which is a quite different situation.
If this is confusing, consider rolling a standard 6-sided die.
The probability, in a single roll, of rolling a number, say $r > 4$ is easy
to compute: it's 2 sides out of six, or $p(r>4) = 1/3$.
If, on the other hand, you roll 10 dice and choose the largest number among
them, the probability that it will be greater than 4 is {\it far} larger
than $1/3$; it is $p({\rm max}_{10}(r) > 4) = 1 - (1 - 1/3)^{10} \approx 0.98$.
The case for the periodogram is analogous: the probability of seeing a
spurious peak at any single location (\eq{LS-prob-z}) is relatively small,
but the probability of seeing a single spurious peak among a large number
of frequencies is much higher.

@@

A more sophisticated treatment of the problem is that of \citet{Baluev2008},
who derived an analytic result based on theory of extreme values
for stochastic processes.
For the PSD-normalized periodogram, \citet{Baluev2008} showed that the following
formula for the FAP provides a close upper-limit even in the case of
highly structured window functions:
\begin{equation}
  FAP(z) \approx 1 - P_{single}(z)e^{-\tau(z)}
  \eqlabel{FAP-baluev}
\end{equation}
where, for the normalized periodogram (\eq{lomb-scargle-normalized}),
\begin{equation}
  \tau(z) \approx W (1 - z)^{(N - 4)/2}\sqrt{z}
\end{equation}
and $W = f_{max} \sqrt{4\pi\ {\rm var}(t)}$ is an effective width of the
observing window in units of the maximum frequency chosen for the analysis.
This result is not an exact measure of the FAP, but rather an
upper limit that is valid for alias-free periodograms, {i.e.} cases
where the window function has very little structure,
but which holds quite well even in the case of more realistic survey windows.

\subsubsubsection{Bootstrap Method}
In the absence of a true analytic solution to the False Alarm Probability, we
can turn to computational methods such as the bootstrap.
The bootstrap method is a technique in which the statistic in question is
computed repeatedly on many random resamplings of the data in order to
approximate the distribution of that statistic
\citep[see][for a useful general discussion of this technique]{ICVG2014}.
For the periodogram, in each resampling we keep the temporal coordinates
the same, draw observations randomly with replacement from the observed
values, and then compute the maximum of the resulting periodogram.
For enough resamplings, the distribution of these maxima will approximate
the true distribution for the case with no periodic signal present.
The bootstrap produces the most robust estimate of the FAP because it makes
few assumptions about the form of the periodogram distribution,
and fully accounts for survey window effects.

@@

Unfortunately, there is no silver bullet for answering
these broader, more relevant questions of uncertainty of Lomb-Scargle results.
Perhaps the most fruitful path toward understanding of such effects for a
particular set of observations---with particular noise characteristics and a
particular observing window---is via simulated data injected into the detection
pipeline.
In fact, this type of simulation is a vital component of most---we might even
venture all---current and future time-domain surveys that seek to
detect, characterize, and catalog periodic objects,
regardless of the particular approach used to identify periodicity
\citep[see, e.g.][etc.]{opsim1, opsim2, Ivezic08LSST, Sesar2010,
  Oluseyi12, McQuillan2012, Drake2014}

@@

\subsubsection{PSD normalization}
When considering the periodogram from the Fourier perspective,
it is useful to normalize the periodogram such that in the special case
of equal spaced data, it recovers the standard Fourier power spectrum.
This is the normalization used in \eq{lomb-scargle-periodogram} and the
equivalent least-squares expression seen in \eq{lomb-scargle-chi2}:
\begin{equation}
  P(f) = \frac{1}{2}\big[\hat{\chi}^2_0 - \hat{\chi}^2(f)\big]
  \eqlabel{lomb-scargle-chi2-2}
\end{equation}
For equally-spaced data, this periodogram is identical to the standard
definition based on the fast Fourier transform:
\begin{equation}
  P(f) = \frac{1}{N} \left| FFT(y_n) \right|^2
\end{equation}
in particular, this means that the units of the periodogram are $unit(y)^2$,
and can be interpreted as squared-amplitudes of the Fourier component at
each frequency.
Note that this changes if data uncertainties are incorporated into
the periodogram as in \sect{extensions-observational-noise},
the periodogram in this normalization becomes unitless: it is essentially
a measure of periodic content in signal-to-noise ratio
rather than in signal itself.

\subsubsection{Least-squares normalization}
In the least-squares view of the periodogram, the periodogram is interpreted
as an inverse measure of the goodness-of-fit for a model.
When we express the periodogram as a function of $\chi^2$ model fits as in
\eq{lomb-scargle-chi2-2}, it becomes clear that the periodogram has a
mathematical maximum value: if the sinusoidal model {\it perfectly} fits
the data at some frequency $f_0$, then $\hat{\chi}^2(f_0) = 0$ and the
periodogram is maximized at a value of $\hat{\chi}^2_0 / 2$.
On the other end, it is mathematically impossible for a sinusoidal model to
do worse than the simple constant, non-varying fit reflected in
$\hat{\chi}^2_0$, and so the minimum value of \eq{lomb-scargle-chi2-2} is
exactly zero.

@@

This paper has been a conceptual tour of the Lomb-Scargle periodogram, from
its roots in Fourier analysis, to its equivalence with special cases of
periodic analysis based on least squares model fitting and Bayesian
analysis.
From this conceptual understanding, we considered a list of challenges
and issues to be considered when applying the method in practice.
We will finish here with a brief summary of these practical
recommendations, a brief discussion of other methods for detecting
periodicity, and a somewhat opinionated
post-script for further thought.

@@

\subsection{Postscript: Why Lomb-Scargle?}
\sectlabel{postscript}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers tend to gravitate toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.

As discussed in \sect{schuster-to-lomb-scargle}, the periodogram has two
benefits over the Schuster approach:
the noise distribution at each individual frequency is the sum
of two Gaussians, and the result is equivalent to a periodogram derived
from a least squares analysis.
But somehow along the way, a mythology seems to have developed surrounding
the efficiency and efficacy of the Lomb-Scargle approach. In particular,
it's common to see statements or implications along the lines of the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with little extra
    computation \citep[see the discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant question of
    maximum peak heights across multiple frequencies,
    which must be either approximated
    or computed by bootstrap resampling (see \sect{uncertainties})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions common to most astronomical
    applications, the Lomb-Scargle
    periodogram has the same window-driven issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: Bayesian
    analysis shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding the frequency of a generic, potentially
    non-sinusoidal signal (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths set aside, what is the practical advantage of Lomb-Scargle
over a Schuster approach?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

The advantage of analytic statistics for Lomb-Scargle evaporates in light of the
need to account for multiple frequencies, so the only advantage left is the
correspondance to least squares and Bayesian models, and in particular the
ability to generalize to more complicated models where appropriate---but in
this case you're not really using Lomb-Scargle at all,
but rather a generative Bayesian model for your
data based on some strong prior information about the form of your signal.
The equivalence of Lomb-Scargle to a Bayesian sinusoidal model is perhaps
interesting trivia, but not itself a reason to use that model if your data
is not known {\it a priori} to be sinusoidal---it could even be construed as
an argument {\it against} Lomb-Scargle in the general case where the assumption
of a sinusoid is not well-founded.

Conversely, if you replace your Lomb-Scargle approach with a Schuster approach,
what you gain is the ability to reason quantitatively about the effects of the
survey window function on the resulting periodogram (see \sect{CLEAN}).
While the deconvolution problem is ill-posed, there is no reason to assume
this is a fatal defect: ill-posed linear models are solved routinely in
other areas of computational research, particularly by using priors or
various forms of regularization.
In any case, I would contend that there is ample room for practitioners to
question the prevailing folk wisdom of the advantage of Lomb-Scargle
over approaches based directly on the Fourier transform and Schuster
periodogram.

>>>>

We are generally not interested in the distribution of one particular
randomly-chosen frequency,
but rather the distribution of {\it the highest peak} of the
periodogram, which is a quite different situation.

By analogy, consider rolling a standard 6-sided die.
The probability, in a single roll, of rolling a number, say $r > 4$ is easy
to compute: it's 2 sides out of six, or $p(r>4) = 1/3$.
If, on the other hand, you roll 10 dice and choose the largest number among
them, the probability that it will be greater than 4 is {\it far} larger
than $1/3$; it is $p({\rm max}_{10}(r) > 4) = 1 - (1 - 1/3)^{10} \approx 0.98$.
The case for the periodogram is analogous: the probability of seeing a
spurious peak at any single location (\eq{LS-prob-z}) is relatively small,
but the probability of seeing a single spurious peak among a large number
of frequencies is much higher.

@@

A more sophisticated treatment of the problem is that of \citet{Baluev2008},
who derived an analytic result based on theory of extreme values
for stochastic processes.
For the standard periodogram in \eq{lomb-scargle-periodogram},
\citet{Baluev2008} showed that the following
formula for the FAP provides a close upper-limit even in the case of
highly structured window functions:
\begin{equation}
  FAP(z) \approx 1 - P_{single}(z)e^{-\tau(z)}
  \eqlabel{FAP-baluev}
\end{equation}
where, for the normalized periodogram (\eq{lomb-scargle-normalized}),
\begin{equation}
  \tau(z) \approx W (1 - z)^{(N - 4)/2}\sqrt{z}
\end{equation}
and $W = f_{max} \sqrt{4\pi\ {\rm var}(t)}$ is an effective width of the
observing window in units of the maximum frequency chosen for the analysis.
This result is not an exact measure of the FAP, but rather an
upper limit that is valid for alias-free periodograms, {i.e.} cases
where the window function has very little structure,
but which holds quite well even in the case of more realistic survey windows.

\subsubsubsection{Bootstrap Method}
In the absence of a true analytic solution to the false alarm probability,
we can turn to computational methods such as the bootstrap.
The bootstrap method is a technique in which the statistic in question is
computed repeatedly on many random resamplings of the data in order to
approximate the distribution of that statistic
\citep[see][for a useful general discussion of this technique]{ICVG2014}.
For the periodogram, in each resampling we keep the temporal coordinates
the same, draw observations randomly with replacement from the observed
values, and then compute the maximum of the resulting periodogram.
For enough resamplings, the distribution of these maxima will approximate
the true distribution for the case with no periodic signal present.
The bootstrap produces the most robust estimate of the FAP because it makes
few assumptions about the form of the periodogram distribution,
and fully accounts for survey window effects.

@@

Unfortunately, there is no silver bullet for answering
these broader, more relevant questions of uncertainty of Lomb-Scargle results.
Perhaps the most fruitful path toward understanding of such effects for a
particular set of observations---with particular noise characteristics and a
particular observing window---is via simulated data injected into the detection
pipeline.
In fact, this type of simulation is a vital component of most (if not all)
current and future time-domain surveys that seek to
detect, characterize, and catalog periodic objects,
regardless of the particular approach used to identify periodicity
\citep[see, e.g.][etc.]{opsim1, opsim2, Ivezic08LSST, Sesar2010,
  Oluseyi12, McQuillan2012, Drake2014}

@@

\subsubsection{PSD normalization}
When considering the periodogram from the Fourier perspective,
it is useful to normalize the periodogram such that in the special case
of equal spaced data, it recovers the standard Fourier power spectrum.
This is the normalization used in \eq{lomb-scargle-periodogram} and the
equivalent least-squares expression seen in \eq{lomb-scargle-chi2}:
\begin{equation}
  P(f) = \frac{1}{2}\big[\hat{\chi}^2_0 - \hat{\chi}^2(f)\big]
  \eqlabel{lomb-scargle-chi2-2}
\end{equation}
For equally-spaced data, this periodogram is identical to the standard
definition based on the fast Fourier transform:
\begin{equation}
  P(f) = \frac{1}{N} \left| FFT(y_n) \right|^2
\end{equation}
in particular, this means that the units of the periodogram are $unit(y)^2$,
and can be interpreted as squared-amplitudes of the Fourier component at
each frequency.
Note, however, that the units change if data uncertainties are incorporated
into the periodogram as in \sect{extensions-observational-noise};
the periodogram in this normalization becomes unitless: it is essentially
a measure of periodic content in signal-to-noise ratio
rather than in signal itself.

\subsubsection{Least-squares normalization}
In the least-squares view of the periodogram, the periodogram is interpreted
as an inverse measure of the goodness-of-fit for a model.
When we express the periodogram as a function of $\chi^2$ model fits as in
\eq{lomb-scargle-chi2-2}, it becomes clear that the periodogram has a
mathematical maximum value: if the sinusoidal model {\it perfectly} fits
the data at some frequency $f_0$, then $\hat{\chi}^2(f_0) = 0$ and the
periodogram is maximized at a value of $\hat{\chi}^2_0 / 2$.
On the other end, it is mathematically impossible for a best-fit sinusoidal
model at any frequency to do worse than the simple constant, non-varying
fit reflected in $\hat{\chi}^2_0$, and so the minimum value of
\eq{lomb-scargle-chi2-2} is exactly zero.

@@

This paper has been a conceptual tour of the Lomb-Scargle periodogram, from
its roots in Fourier analysis, to its equivalence with special cases of
periodic analysis based on least squares model fitting and Bayesian
analysis.
From this conceptual understanding, we considered a list of challenges
and issues to be considered when applying the method in practice.
We will finish here with a brief summary of these practical
recommendations, along with a somewhat opinionated
post-script for further thought.

@@

\subsection{Postscript: Why Lomb-Scargle?}
\sectlabel{postscript}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers tend to gravitate toward the
Lomb-Scargle approach rather than the (in many ways simpler) classical
periodogram.

As discussed in \sect{schuster-to-lomb-scargle}, the Lomb-Scargle approach has
two distinct benefits over the classical periodogram:
the noise distribution at each individual frequency is chi-square distributed
under the null hypothesis,
and the result is equivalent to a periodogram derived
from a least squares analysis.
But somehow along the way, a mythology seems to have developed surrounding
the efficiency and efficacy of the Lomb-Scargle approach. In particular,
it's common to see statements or implications along the lines of the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the classical periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the classical periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with little extra
    computation \citep[see the discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant question of
    maximum peak heights across multiple frequencies,
    which must be either approximated
    or computed by bootstrap resampling (see \sect{uncertainties})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions common to most astronomical
    applications, the Lomb-Scargle
    periodogram has the same window-driven issues as the classical
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: Bayesian
    analysis shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding the frequency of a generic, potentially
    non-sinusoidal signal (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths set aside, what is the practical advantage of Lomb-Scargle
over a classical periodogram?
What would we lose if we instead used a simple Fourier periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

The advantage of analytic statistics for Lomb-Scargle evaporates in light of the
need to account for multiple frequencies, so the only advantage left is the
correspondence to least squares and Bayesian models, and in particular the
ability to generalize to more complicated models where appropriate---but in
this case you're not really using Lomb-Scargle at all,
but rather a generative Bayesian model for your
data based on some strong prior information about the form of your signal.
The equivalence of Lomb-Scargle to a Bayesian sinusoidal model is perhaps
an interesting bit of trivia, but not itself a reason to use that model if your
data is not known {\it a priori} to be sinusoidal---it could even be construed as
an argument {\it against} Lomb-Scargle in the general case where the assumption
of a sinusoid is not well-founded.

Conversely, if you replace your Lomb-Scargle approach with a classical periodogram,
what you gain is the ability to reason quantitatively about the effects of the
survey window function on the resulting periodogram (see \sect{CLEAN}).
While the deconvolution problem is ill-posed, there is no reason to assume
this is a fatal defect: ill-posed linear models are solved routinely in
other areas of computational research, particularly by using priors or
various forms of regularization.
In any case, I would contend that there is ample room for practitioners to
question the prevailing folk wisdom of the advantage of Lomb-Scargle
over approaches based directly on the Fourier transform and classical
periodogram.
